---
title: "Ch1 - Data Visualization with ggplot2"
author: "L. Pandori (adapted from Wickham & Grolemund)"
date: "4/6/2020"
output: html_document
---
## Packages and Data Prep
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load library
library(tidyverse)

# Load dataset
mpg <- ggplot2::mpg
```

## First steps
 displ = engine size (L)
 hwy = highway fuel efficiency (mpg)

```{r}
ggplot(data = mpg) +
  geom_point (mapping = aes (x = displ, y= hwy))
```

 Template:
 ggplot(data = <DATA>) + 
 <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
 
Exercises
1. If you run ggplot(data = mpg), it shows an empty gray plot

```{r}
ggplot(data = mpg)
```

2. There are 11 columns and 234 rows in the dataframe mpg

```{r}
ncol(mpg)
nrow(mpg)
```

3. The drv variable describes the type of drive train

4. Make a scatterplot of hwy (highway mpg) vs cyl (# cylinders)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = hwy, y = cyl))
```

5. Make a scatterplot of class (car type) vs drv (drive train) - it is not useful bc these are categorical variables

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv))
```

## Aesthetic mappings
Aes = aesthetic = visual property of objects in a plot
Used to add 3rd+ variable to 2D plots with:
 * size (size = )
 * shape (shape = ) - only works w/ <= 6 categories
 * color (color = )
 * transparency (alpha = )

Ex) can add class of car to scatterplot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

To set aes properties manually, put outside aes argument
Ex) ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), **color = 'blue'**)

Exercises
1. The points are not blue because the color argument is inside the aes argument

2. Which variables for mpg are continuous vs categorical? How do you know?
categorical: manufacturer, model, year, drv, fl, class
continuous: displ, cyl, cty, hwy

3. Map a continuous variale to color, size, and shape

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = class))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, size = cyl))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, shape = drv))
```

For arguemnts that work with both cat/continuous vars, continuous have a gradient and categorical variables have a legend

4. If you map the same variable to multiple aesthetics...

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = drv, shape = drv))
```

Nothing terrible happens...you just get redundant labeling

5. The 'stroke' aesthetic changes point border thickness -- but can only be used for pch with fill and size arguments

6. If you map an aesthetic to something other than a variable name, 
the points will be colored by whether that statement is true or false

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = displ < 5))
```

## Facets
Splits data into subplots by levels of a factor

For a single factor, use facet_wrap
plot + facet_wrap (~ factor, nrow = XX)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy)) + 
  facet_wrap (~ class, nrow = 3)
```

For multiple factors, use facet_grid
plot + facet_grid(factor1 ~ factor 2)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy)) + 
  facet_grid (fl ~ class)
```

Exercises
1. If you use facets on continuous variables, ech unique value for that variable will create a facet -- which results in a strange plot

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy)) + 
  facet_wrap (~ displ)
```

2. Empty cells in the plot below indicate that there are no data points where those cells intercept (for ecample, no cars that have 5 cylinders that are 4wd)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes (x = drv, y = cyl)) + 
  facet_grid(drv ~ cyl)
```

3. Using (. ~ XX) in facet_grid makes subsets vertical (x-axis)
   Using (XX ~ .) in facet_grid makes subsets horizontal (y-axis)
   
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes (x = displ, y = hwy)) + 
  facet_grid (. ~ drv)

ggplot(data = mpg) + 
  geom_point(mapping = aes (x = displ, y = hwy)) + 
  facet_grid(drv ~ .)
```  

4. For this dataset, I prefer aesthetic arguments over facets. However, with a larger dataset, where overlapping points become more of an issue, I would prefer facets. 

5. Facet_wrap makes 1d sequence of panels 2d. The arguments nrow and ncol specify the # of rows and columns in the 2d display. 

## Geometric Objects
geom = geometric object to represent data
Layer geoms to add multiple representations of data to a plot

To add layers with same x and y coordinates, specify in ggplot() argument rather than geom() argument -- see example below

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes (x = displ, y = hwy)) + 
  geom_smooth(mapping = aes( x = displ, y = hwy))

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```  

If you place mapping arguments in geom functions, it will work for only that function

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
``` 

Exercises
1. Geom for: 
 * Line chart - geom_line
 * Boxplot - geom_boxplot
 * Histogram - geom_histogram
 * Area chart - geom_freqpoly

2. Theoretical

3. 'show.legend' when TRUE shows a legend for lines to R of plot, if FALSE no legend displayed

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(color = drv),
              show.legend = TRUE)
```

4. The 'se' argument to geom_smooth adds a CI (or not).

5. These two plots will not look different (tested in code -- not incl here)

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
# vs
ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

6. Re-create R code necessary for following graphs (see text pg 21)
```{r}
# plots recreated L -> R, top -> bottom
# plot A (top L)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

# plot B
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(group = drv), se = FALSE) + 
  geom_point()

# plot C
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(color = drv), se = FALSE) + 
  geom_point(mapping = aes(color = drv))

# plot D
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(se = FALSE) + 
  geom_point(mapping = aes(color = drv))

# plot E
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(linetype = drv), se = FALSE) + 
  geom_point(mapping = aes(color = drv))

# plot F
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = drv))
```

## Statistical transformations
Using diamonds dataset in ggplot2

stat = statistical transformation = charts which plot summary data
includes: bar, hist, freq 

Smoothers fit a model to data + plot predictions

Can use geoms and stats interchangably
Ex) 
```{r}
ggplot(data = diamonds, mapping = aes(x = cut)) +
  geom_bar()
# vs
ggplot(data = diamonds, mapping = aes(x = cut)) +
  stat_count()
```

Sometimes you won't want to use the default stat for a geom...
 * Bar heights already in the data frame (stat = 'identity
 )
 
```{r}
# give it a made-up dataset
demo <- tribble(
  ~a, ~b,
  'bar1', 10,
  'bar2', 15,
  'bar3', 30)

# plot it - we already have bar heights in 'b'
ggplot(data = demo) + 
  geom_bar(mapping = aes(x = a, y = b), stat = 'identity')
```

 * Plot proportions (y = ..prop..)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = ..prop.., group = 1)) + 
  geom_bar()
```

 * Any summar stats of your choice (stat_summary)
   For full list of stats use ?stat_bin (or ggplot2 cheat sheet)
  
```{r}
ggplot(data = diamonds) + 
  stat_summary(mapping = aes(x = cut, y = depth),
  fun.ymin = min,
  fun.ymax = max,
  fun.y = median)
```

Exercises 
1. The default geom for stat_summary is range

2. Geom_bar differs from geom_stat bc geom_bar uses count data, geom_col uses raw data

3. No

4. stat_smooth() computes summary data as a method (function) = which can be lm, glm, gam, or loess

5. Setting group = 1 means the sum of the proportions = 1. 

```{r}
# without group = 1 argument 
ggplot(data = diamonds) + 
  geom_bar( mapping = aes(x = cut, y = ..prop..))

# with group = 1 argument
ggplot(data = diamonds) + 
  geom_bar( mapping = aes(x = cut, y = ..prop.., group = 1))
```

## Position Adjustments
Bar plots -- in aes argument...
color = color outside of bars
fill = fill bars (automatically stacked if you give it a third variable -- not in the x or y argument)

```{r}
# color argument
ggplot(data = diamonds) + 
  geom_bar( mapping = aes(x = cut, color = cut))

# fill argument
ggplot(data = diamonds) + 
  geom_bar( mapping = aes(x = cut, fill = cut))

# auto-stacked bar when 3rd variable added
ggplot(data = diamonds) + 
  geom_bar( mapping = aes(x = cut, fill = clarity))
```

Auto-stacked bar is indicated by position argument
For other position arguments w/in geom_bar...
  * position = 'identity' places each object where it should be in space
  
```{r}
# color argument
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = 'identity')
```

  * position = 'fill' stacks, but all bars same height
  
```{r}
# color argument
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(position = 'fill')
```  
  
  * position = 'dodge' groups bars by category
  
```{r}
# color argument
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(position = 'dodge')
```   
  
  * position = 'jitter' -- not for bars, but jitters points in scatterplots
  
Exercises 

1. This plot could be improved by jittering points -- there are likely multiple observations @ each x,y

2. In geom_jitter, the amount of jitter-y-ness can be controled by the width and height of jitteredness (width = and height = ) and if needed, to be reproducible (seed = XX)

3. Jittering (geom_jitter) displaces points to make each observation more visible, while geom_count scales point area to abundance. 

4. The default position argument for geom_boxplot is 'dodge2'. It is demonstrated in this plot: 

```{r}
# color argument
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_boxplot(position = 'dodge')
``` 

## Coordinate systems
default = cartesian (x, y)

coord_flip - flips x and y axes (horizontal boxplots, long labels)
coord_quickmap - sets aspect ratio correctly for maps 
coord_polar - polar coodinates (maps bars to round areas - rose plot)

Exercises
1. Turn a stacked bar into a pie chart using coord_polar

```{r}
# stacked bar
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

# polar coordinates
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity)) + 
  coord_polar()
```

2. What does labs do?
labs(title = waiver(), - chart title
    subtitle = waiver(), - chart sub-title
    caption = waiver(), - caption below figure
    tag = waiver()) - top L plot label

xlab () - x-axis label
ylab () - y-axis label
ggtitle () - main chart title


3. Coord_map and coord_quickmap differ b/c coord_map doesn't preserve straight lines while coord_quickmap does

4. The plot tells us there is a + linear relationship between city and hwy mpg. 
coord_fixed makes 1 unit on x axis = 1 unit on y axis (use when same units)
geom_abline adds a y=mx+b lm to the model and plots the result

