---
title: "Ch3_DataTransformation_dplyr"
author: "L. Pandori (adapted from Wickham & Grolemund)"
date: "4/7/2020"
output: html_document
---
## Intro & Prerequisites

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load library
library(tidyverse) # ggplot2 and dplyr
library(nycflights13) #dataset
flights <- flights
# ?flights # for more info
```

Look at warning messages...if dplyr has overwritten a function from base R, you can still use it by specifying *package::function*

NYC Flights 13 - flight info for 336k flights from NYC in 2013

Data types - 
 * int = integers
 * dbl = double/real numbers
 * chr = character vectors/strings
 * dttm = date-time
 * lgl = logical (vectors w/ only TRUE or FALSE)
 * fctr = factor (categorical var with fixed possible values)
 * date = date

dplyr basics - 

Format is *function(df, what to do)*, result in new df
Fns don't change df...to save as new use *newdf <- fn(olddf, XX)*

To save the 
 * filter() = pick obs by values
 * arrange() = reorder rows
 * select() = pick variables by names
 * mutate() = create new variables w f(n) of existing variables
 * summarize() = collapse many values to summarry

All can be used with group_by(), which makes functions operate on each group

## filter() = pick obs by varibles

Comparisons: == (equals), >, <, >=, <=, != (not equal)
*note* when doing arithmetic, use near() instead of == 

Ex) 
```{r}
sqrt(2)^2 == 2
near(sqrt(2)^2, 2)
```

Logical operators: & (and), | (or), ! (not)
*note* you can't use these operators without repeating arguments. As an alternative, use %in%

```{r}
filter(flights, month == 11|12) #doesn't work (see month = 1 in row 1)
filter(flights, month == 11 | month == 12) # works
filter(flights, month %in% c(11,12)) # works, more efficient
```

DeMorgan's Law
  * !(x & y) = !x | !y
  * !(x | y) = !x & !y
  
Missing Values
NAs = missing values 

To determine if values is missing, use *is.na()*

Exercises: 
1. Find all fights that meet criteria listed in book

```{r}
# a
head(filter(flights, arr_delay >= 60))
# b
head(filter(flights, dest %in% c('IAH', 'HOU')))
# c
head(filter(flights, carrier %in% c('UA', 'AA', 'DL')))
# d
head(filter(flights, month %in% c('7', '8', '9')))
# e
head(filter(flights, dep_delay <= 0 & arr_delay > 120))
# f 
head(filter(flights, dep_delay >= 60 & arr_delay <= dep_delay - 30))
# g
head(filter(flights, dep_time == 2400 | dep_time <= 600))
```

2. between() operator - (x, left, right)
can be used to simplify observations like #d in ex above
filter(flights, between(month, 7,9))

3. 8255 flights have a missing dept_time. They might represent cancelled flights, bc they are missing an arrival time & flight duration.

```{r}
nrow(filter(flights, is.na(dep_time)))
```

4. N^0 and N*0 are not missing because any number ^0 = 1 and any number x 0 = 0. 

## arrange() - to order rows
Works like filter() arguments, but just changes order
If > 1 argument, additional arugments break ties

Exercises
1. Sort all missing values to start of dataframe

```{r}
head(arrange(flights, desc(is.na(dep_time), dep_time)))

```

2. Find most delayed flight and earliest departing flight

```{r}
# find most delayed flight (1301 min late)
head(arrange(flights, desc(dep_delay)))
# find least delayed flight (43 min early)
head(arrange(flights, dep_delay))
```

3. Fastest flight = 20 mins

```{r}
head(arrange(flights, air_time))
```

4. Longest travel distance = 4983 miles
Shortest = 17 miles

```{r}
# shortest distance
head(arrange(flights, distance))
# longest distance
head(arrange(flights, desc(distance)))
```

## select()
Narrow df to columns of interest
 * -(list) to exclude
 * () to include
 * (:) to include between

Helper functions: 
 * starts_with ('abc')
 * ends_with ('abc')
 * contains ('abc')
 * matches (*regular expression - see ch11*)

To re-name variables, use rename() -- not select()
  Preserves other variables as well 
  ex) rename(flights, tail_num = tailnum)
  
To move vars of interest to L side of df...
  select(*df*, *v1*, *v2* everything())
  
Exercises: 

1. Multiple ways to get the same columns

```{r}
# way 1
head(select(flights, dep_time, dep_delay, arr_time, arr_delay))
# way 2
head(select(flights, starts_with('dep_'), starts_with('arr_')))
```

2. If var name is included more than once, it will only be included once

3. one_of() function is retired
any_of() - selects all that match (OK to have ones that may not exist)
all_of() - selects any that match 

4. contains() is not case sensitive

## mutate() adds new vars to R side of df
If you only want to keep new vars, use transmute()

Creation functions (input and output are vectors)
 * Arithmetic operators = +, -, *, /, ^
 * Modular arithmetic
    %/% - integer division
    %% - remainder
 * Logs - log(*ln*), log2(*base2*), log10(*base10*)
 * Offsets - lead(*leads by x*), lag(*lags by x*)
 * Cumulative/rolling agregates
    cumsum(), cumprod(), cummin(), cummax(), cummean()
    *For rolling aggregates, try RcppRoll package*
 * Logical comparisons: >, <=, >=, <, != etc.
 * Ranking
    # smallest to largest min_rank(*df$col*)
    # largest to smallest min_rank(desc(*df$col*))

Exercises

1. convert dep_time and arr_time to mins since midnight #nope

```{r}
flights$dep_time_mm <- mutate(flights, dep_time %/% 100 * 60 + dep_time %% 100, dep_time)
```
  
## summarize() - grouped summaries
group_by() to specify groups to summarize obs in

Functions: 
 * Measures of location: mean(x), median(x)
 * Measures of spread: sd(x), IQR(x), mad(x) - median absolute deviation
 * Measures of rank: min(x), quantile(x, *0.25*), max(x)
 * Measures of position: first(x), nth(x,*#*), last(x)
 * Number of unique observations: n_distinct()

## %>% = pipe, combines operations
Can combine, for example
1. Select/group data of interest *flights %>% group_by(dest) %>%*
2. Calculate summary *summarize(count = n(), dist = mean(distance, na.rm = TRUE) %>%*
3. Filter to remove outliers *filter(count > 20, deat != 'HNL')*

*na.rm = TRUE* removes missing values before calculation
*count(n())* or *sum(!is.na(x))* give sample size

## Grouping multiple variables
*group_by()*
to ungroup, use *ungroup()*

Exercises: 

1. 

```{r}
# 15 min early 50% of the time, 15 min late 50% of the time
min15 <- flights %>%
          group_by()

# always 10 min late
late10 <- filter(flights, arr_delay == 10)
unique(late10$arr_delay)

# 30 min early 50% of the time, 30 min late 50% of the time


# 99% on time, 1% 2 hrs late
```

2. Same output, without using count()

```{r}
# original code
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))
not_cancelled %>% count(dest)

# new way w/o count()
not_cancelled %>% 
  group_by(dest) %>%
  summarize(n = length(arr_time))

# original code
not_cancelled %>% count(tailnum, wt = distance)

# new way w/o count()
not_cancelled %>%
  group_by(tailnum) %>%
  summarize(n = sum(distance))
```

3. In defining cancelled flights, dep_delay is most important.

4. Look at # of cancelled flights per day -- pattern?

```{r}
# find # of cancelled flights per day
cancelled <- flights %>%
  filter(is.na(arr_delay), is.na(dep_delay)) %>%
  group_by(date(time_hour)) %>%
  summarize(n = length(arr_delay))
# look at top 10 observations
# no distinguishable pattern (just from eyeball)
```

5. Which carrier has worst delays? - of course, it's Frontier

```{r}
# only group by carrier
carrierdelay <- flights %>% 
  group_by(carrier) %>%
  summarize(mean_delay = mean(arr_delay, na.rm = TRUE),
            sd_delay = sd(arr_delay, na.rm = TRUE))
```

6. For each plane, count # of obs before first delay > 1 hr
!!! I can't figure this one out yet !!!
```{r}
obs <- flights %>%
  group_by(tailnum) %>% # group by plane
  arrange(tailnum, year, month, day) %>% # sort by yr, mo, day
  summarize(days = rank(arr_delay > 60))
```

7. The sort argument to count() can replace count followed by arrange

## Group mutates and filters



