---
title: "Ch5_Exploratory_Data_Analysis"
author: "L. Pandori (adapted from Wickham & Grolemund)"
date: "4/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro
Exploratory data analysis = use vis & transformation to explore data in a systematic way with goals: 
 * generate q's
 * search for answers
 * use what you learn to refine q's or generate new q's
 
Possible guiding q's
 * what variation occurs?
 * what covariation occurs?

Definitions: 
1. variable - measured quantity/quality/property
2. value - state of var when measured
3. obs/case - set of measurements made under similar conditions
4. tabular data - tidy if each value in own cell, each var in column, each obs in row

## Variation

Visualizing distributions: 
  - barplot if categorical (geom_bar)
  - histogram if continuous (geom_histogram)
  (use geom freq_poly for multiple histograms on same plot)

Think about 
  * which values most common & why
  * which rare & why
  * notice unusual patterns & why

Clusters of similar obs indicate subgroups in data 

What about outliers? (make bars thick to see more clearly)
 * repeat analysis w/ and w/o outliers 

Exercises: 
1. explore x (length, mm), y (width), z (depth) variables of diamonds
 * they are continuous
```{r}
library(tidyverse)
ggplot(data = diamonds) + 
  geom_histogram(mapping = aes(x = x), fill = 'blue', alpha = 0.5) + 
  geom_histogram(mapping = aes(x = y), fill = 'green', alpha = 0.5) +
  geom_histogram(mapping = aes(x = z), fill = 'yellow', alpha = 0.5) + 
  coord_cartesian(xlim = c(0,10)) + 
  xlab(label = 'length, width, depth (mm)')
  theme_bw()
```

2. explore price
```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_histogram(binwidth = 1000) +
  theme_bw()
```

3. how many diamonds are 0.99 carat? 23
how many 1 carat? 1558

```{r}
ggplot(data = diamonds, mapping = aes(x = carat)) + 
  geom_histogram(binwidth = 0.01) +
  theme_bw()

nrow(diamonds %>% filter(carat == 1))
nrow(diamonds %>% filter(carat == 0.99))
```

4. exploratory question -- done in console

## Dealing with missing/unusual values
2 options: 
 * drop entire row with filter (not recommended)
 * replace with NAs (recommended) with mutate
 ex) df2 <- df1 %>% mutate(var = ifelse(condition, NA, var))
 
 Keep in mind what missing values mean (like cancelled flights)
 
 Exercises: 
 1. NAs are not counted in histogram, but make own column in barplot
 2. na.rm = TRUE discards NAs from calculation of summary stats (in mean() and sum())
 
## Covariation
Describes variation b/w vars

Categorical x continuous - look at boxplots
  * IQR = length of box (25%tile to 75%tile), middle is median (50%tile)
  * whiskers are 1.5 x IQR
  * outliers as points

Can re-order boxplots using mapping = aes(x = reorder(x = x, y = y, FUN = median), y = y)

Exercises - skipped, I feel practiced in boxplot wrangling

## 2 Categorical Variables
Count # of obs for each combination - geom_count() works well
Covariation appears as strong correlation b/w x and y values

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = color)) + 
  geom_count() +
  theme_bw()

# or

ggplot(data = diamonds %>% count(color, cut), mapping = aes(x = color, y = cut)) + 
  geom_tile(mapping = aes(fill = n))
```

## 2 Continuous Variables
Scatterplot w/ geom_point()
Covariation indicated by linear relationship b/w points

## Patterns & Models
Questions to ask about patterns: 
  * Due to chance?
  * How to describe relationship implied?
  * How strong?
  * Do other vars affect relationship?
  * Does relationship change if viewed by group?


