---
title: "Ch789_Input_Tidying"
author: "L. Pandori (adapted from Wickham & Grolemund)"
date: "4/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ch 7 - Tibbles
Use over df b/c: 
 * when called, gives 10x10 output
 * when subsetted, always returns tibble

```{r}
# more info
vignette('tibble')
```

## Ch 8 - Data import w/ readr
name <- read_csv('file.csv')
 * add *na = 'navalue'* to specify NAs

# Parsing a vector - tell R how to treat a column (what type of data?)
parse*_(character vector to parse, na = 'navalue')
 * logical, integer
 * double (strict numeric)
 * number (flexible numeric) - ignores things before/after # (ex - %$)
 * character - gets complicated w/ older files (see txt pg 133)
 * factor - cat vars w known set of values
 * datetime, date, time - specify format (see below)
    year - %y (2 digit), %Y (4 digit)
    month - %m (2 digits), %b (abbr name), %B (full name)
    day - %d (2 digits), %e (opt leading space)
    time - %H (hour 0-23), %I (hour 0-12, must be used with %p), 
           %p (am/pm), %M (min), %S (integer seconds), 
           %OS (real seconds), %Z (time zone - see pg 254), 
           %z (offset from UTC)

# Parsing a file
readr automatically parses after analyzing first 1k rows, but can be overridden 

name <- read_csv('file.csv',
        col_types = cols(
        x = col_*(),
        y = col_*())
        
?? When do you use parse_*() then? - when data are already in R

Write a file
*write_csv(tibble, 'name.csv')*
for Excel specifically
*write_excel_csv(tibble, 'name.csv')*
to keep important parsing
*write_rds(tibble, 'name.rds')*

## Ch 9 - Tidy Data w Tidyr

Tidy data - long format (easy to work w/)
 * each var in own column
 * each obs in own row
 * each value in own cell 
 
# Spreading (break 1 col into 2) and gathering (make 2 col into 1)
df %>% gather(col1, col2, key = newcolname, value = variabletospread)
spread(df, key = coltosplit, value = variabletogather)

Exercises:

1.
```{r}
library(tidyverse)
stocks <- tibble(
  year = c(2015,2015,2016,2016),
  half = c(1,2,1,2),
  return = c(1.88, 0.59, 0.92, 0.17))

stocks2 <- spread(data = stocks, key = year, value = return)
stocks2 <- gather(data = stocks2, year, return, '2015':'2016')

```

2. Why does this code fail? needs backticks around 1999 and 2000

```{r}
# fail 
table4a %>% 
  gather(1999 , 2000, key = 'year', value = 'cases')
# fix
table4a %>% 
  gather('1999' , '2000', key = 'year', value = 'cases')
```

3. Need to spread the key column - so that age & height are 2 dif columns for each observation (person)

4. Need to gather - so that sex, preg and whatever the male and female obs are are separate

# Separating & Pull
Use separate when one column contains 2 observations
Ex) 258/89 --> 258 |  89
df %>% separate(column, into = c('newcol1', 'newcol2'), 
        sep = 'delimiter',
        convert = T(if want new cols have dif type)/F (presrve type))

Use separate to get first/last values of column
df %>% separate(column, into = c('newcol1', 'newcol2'),
        sep = # (starts with 1 on L and -1 on R))

Use unite to combine multiple cols into single col (inverse of separate)

df %>% unite(newcol, col1, col2)
   where col1 and col2 are cols you want to unite 
   
# Missing Values
explicit - (presence of an absence) values w/ NA or other indication
implicit - (absence of a presence) missing row that fits a pattern

# TB Case Study Dataset



